Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-05-29 output/manuscript.tex
merge with latest text

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
        11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage{lmodern}
\usepackage{iftex}
\usepackage[legalpaper, left=0.5in, right=0.5in, top=1in, bottom=1in]{geometry}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\usepackage[inkscapeformat=pdf,inkscapelatex=false]{svg}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{american}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}

@@

  · \includesvg{images/mastodon.svg}

@@

@@

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.svg}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
PhenoPLIER integrates groups of genes co-expressed in specific cell types (gene modules) with gene-trait and gene-drug associations.
At the bottom-right corner, we show how a drug downregulates two genes (A and C) that are part of the disease-associated Gene module 1.
\textbf{b)} The integration consists of projecting gene-trait/gene-drug associations from PhenomeXcan/LINCS L1000 (bottom) to a latent space based on gene modules (represented by latent variables/LVs) from MultiPLIER (top).
This process generates matrices \(\mathbf{\hat{L}}\) and \(\mathbf{\hat{M}}\), where LVs now describe each drug/trait.
In the middle, we show the three main computational components provided by PhenoPLIER to perform this integration:
1) an LV-based regression model,
2) a clustering framework to learn groups of traits,
and 3) an LV-based drug repurposing approach.
\textbf{c)} LV603, termed as a neutrophil signature in the original MultiPLIER study, was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
Genes that are part of LV603 were expressed in relevant cell types (top).
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

PhenoPLIER is a flexible computational framework that combines gene-trait and gene-drug associations with gene modules expressed in specific contexts (Figure \ref{fig:entire_process}a).
The approach uses a latent representation (with latent variables or LVs representing gene modules) derived from a large gene expression compendium (Figure \ref{fig:entire_process}b, top) to integrate TWAS with drug-induced transcriptional responses (Figure \ref{fig:entire_process}b, bottom) for a joint analysis.
The approach consists in three main components (Figure \ref{fig:entire_process}b, middle, see \protect\hyperlink{sec:methods}{Methods}):
1) an LV-based regression model to compute an association between an LV and a trait,
2) a clustering framework to learn groups of traits with shared transcriptomic properties,
and 3) an LV-based drug repurposing approach that links diseases to potential treatments.
We performed extensive simulations for our regression model (\protect\hyperlink{sm:reg:null_sim}{Supplementary Note 1}) and clustering framework (\protect\hyperlink{sm:clustering:null_sim}{Supplementary Note 2}) to ensure proper calibration and expected results under a model of no association.

@@

\begin{equation}
\begin{split}
Cor(\mathbf{P}_{i}, \mathbf{P}_{j}) & = Cor(\mathbf{T}_{i} \mathbf{V}_{i}^{\top} \mathrm{diag}(\lambda_i)^{-1/2}, \mathbf{T}_{j} \mathbf{V}_{j}^{\top} \mathrm{diag}(\lambda_j)^{-1/2}) \\
& = \mathrm{diag}(\lambda_i)^{-1/2} \mathbf{V}_{i} (\frac{\mathbf{T}_{i}^{\top} \mathbf{T}_{j}}{n-1}) \mathbf{V}_{j}^{\top} \mathrm{diag}(\lambda_j)^{-1/2},
\end{split}
\label{eq:reg:corpp}\end{equation}

@@

\begin{equation}
\begin{split}
\frac{(\mathbf{T}_{i}^{\top} \mathbf{T}_{j})_{kl}}{n-1} & = Cor(\mathbf{t}_k^i, \mathbf{t}_l^j) \\
 & = \frac{ Cov(\mathbf{t}_k, \mathbf{t}_l) } { \sqrt{\widehat{\mathrm{var}}(\mathbf{t}_k) \widehat{\mathrm{var}}(\mathbf{t}_l)} } \\
 & = \frac{ Cov(\sum_{a \in \mathrm{model}_k} w_a^k X_a, \sum_{b \in \mathrm{model}_l} w_b^l X_b) }  {\sqrt{\widehat{\mathrm{var}}(\mathbf{t}_k) \widehat{\mathrm{var}}(\mathbf{t}_l)} } \\
 & = \frac{ \splitfrac{\sum_{a \in \mathrm{model}_k}{b \in \mathrm{model}_l}} w_a^k w_b^l Cov(X_a, X_b)} {\sqrt{\widehat{\mathrm{var}}(\mathbf{t}_k) \widehat{\mathrm{var}}(\mathbf{t}_l)} } \\
        & = \frac{ \splitfrac{\sum_{a \in \mathrm{model}_k}{b \in \mathrm{model}_l}} w_a^k w_b^l \Gamma_{ab}} {\sqrt{\widehat{\mathrm{var}}(\mathbf{t}_k) \widehat{\mathrm{var}}(\mathbf{t}_l)} },
\end{split}
\label{eq:reg:corrgenes}\end{equation}

@@

Note that, since we used the MultiXcan regression model (Equation (\ref{eq:multixcan})), \(\mathbf{R}\) is only an approximation of gene correlations in S-MultiXcan.
As explained before, S-MultiXcan approximates the joint regression parameters in MultiXcan using the marginal regression estimates from S-PrediXcan in (\ref{eq:spredixcan}) with some simplifying assumptions and different genotype covariance matrices.
This complicates the derivation of an S-MultiXcan-specific solution to compute \(\mathbf{R}\).
To account for this, we used a submatrix \(\mathbf{R}_{\ell}\) corresponding to genes that are part of LV \(\ell\) only (top 1\% of genes) instead of the entire matrix \(\mathbf{R}\).
This simplification is conservative since correlations are accounted for top genes only.
Our simulations (\protect\hyperlink{sm:reg:null_sim}{Supplementary Note 1}) show that the model is approximately well-calibrated and can correct for LVs with adjacent and highly correlated genes at the top (e.g., Figure \ref{fig:reg:nulls:qqplot:lv234}).
The model can also detect LVs associated with relevant traits (Figure \ref{fig:lv246} and Table \ref{tbl:sup:phenomexcan_assocs:lv246}) that are replicated in a different cohort (Table \ref{tbl:sup:emerge_assocs:lv246}).

In Equation (\ref{eq:reg:corrgenes}), for each gene, we only considered tissue models present in S-PrediXcan results, as well as SNPs present in GWAS used as input for the TWAS approaches.
This is necessary to obtain more accurate correlations estimates {[}\protect\hyperlink{ref-1FFzCXo1s}{24}{]}.
Therefore, we computed different correlation matrices for PhenomeXcan and eMERGE.
In PhenomeXcan, most of the GWAS (4,049) were obtained from the UK Biobank using the same pipeline and including the same set of SNPs, so a single correlation matrix was used for this set.
For the rest, we used a single correlation matrix for each group of traits that shared the same or most of the SNPs.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.6970}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1136}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0682}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1212}}@{}}
\caption{Significant trait associations of LV246 in PhenomeXcan.
\label{tbl:sup:phenomexcan_assocs:lv246}}\label{tbl:sup:phenomexcan_assocs:lv246}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cases
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
FDR
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cases
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
FDR
\end{minipage} \\
\midrule()
\endhead
Triglycerides NMR & 21,559 & & 1.66e‑26 \\
LDL Cholesterol NMR & 13,527 & & 3.92e‑26 \\
High cholesterol (self-reported) & 361,141 & 43,957 & 1.08e‑24 \\
Cholesterol lowering medication & 193,148 & 24,247 & 4.28e‑24 \\
Treatment/medication code: simvastatin & 361,141 & 40,921 & 2.56e‑19 \\
CH2DB NMR & 24,154 & & 1.05e‑15 \\
Cholesterol lowering medication & 165,340 & 38,057 & 9.58e‑15 \\
Treatment/medication code: atorvastatin & 361,141 & 10,805 & 2.54e‑14 \\
Illnesses of mother: Alzheimer's disease/dementia & 331,041 & 28,507 & 2.76e‑08 \\
Illnesses of father: Alzheimer's disease/dementia & 312,666 & 15,022 & 2.76e‑08 \\
Alzheimers Disease & 54,162 & 17,008 & 1.10e‑07 \\
Non-butter spread type details: Flora Pro-Active or Benecol & 190,094 & 29,048 & 5.63e‑07 \\
Illnesses of siblings: Alzheimer's disease/dementia & 279,062 & 1,609 & 6.16e‑07 \\
Any dementia & 361,194 & 243 & 2.86e‑05 \\
Illnesses of father: None of the above (group 1) & 314,797 & 116,736 & 3.56e‑05 \\
Medication for cholesterol, blood pressure, diabetes, or take exogenous hormones (females) & 193,148 & 133,338 & 1.10e‑04 \\
Treatment/medication code: lipitor 10mg tablet & 361,141 & 2,584 & 1.55e‑04 \\
Treatment/medication code: rosuvastatin & 361,141 & 2,227 & 1.37e‑03 \\
Illnesses of father: Heart disease & 318,570 & 104,110 & 1.89e‑03 \\
Dementia & 361,194 & 157 & 9.58e‑03 \\
Mother still alive & 355,029 & 140,246 & 1.76e‑02 \\
Job SOC coding: Librarians & 91,149 & 1,248 & 3.22e‑02 \\
Alzheimer's disease & 361,194 & 119 & 3.61e‑02 \\
\bottomrule()
\end{longtable}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{mathtools}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\usepackage[inkscapeformat=pdf,inkscapelatex=false]{svg}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{american}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}

@@

  · \includesvg[width=0.16667in,height=0.16667in]{images/mastodon.svg}

@@

  \href{https://orcid.org/0000-0001-7720-6208}{0000-0001-7720-6208}
  \href{https://github.com/bdsullivan}{bdsullivan}
  \href{https://twitter.com/blairdsullivan}{blairdsullivan}

@@

\begin{figure}
\hypertarget{fig:entire_process}{%
\centering
\includesvg[width=1\textwidth,height=\textheight]{images/entire_process/entire_process.svg}
\caption{\textbf{Schematic of the PhenoPLIER framework.}
\textbf{a)} High-level schematic of PhenoPLIER (a gene module-based method) in the context of TWAS (single-gene) and GWAS (single-variant).
In GWAS, we identify variants associated with traits.
In TWAS, first, we identify variants that are associated with gene expression levels (eQTLs); then, prediction models based on eQTLs are used to impute gene expression, which is used to compute gene-trait associations.
Resources such as LINCS L1000 provide information about how a drug perturbs gene expression; at the bottom-right corner, we show how a drug downregulates two genes (A and C).
In PhenoPLIER, these data types are integrated using groups of genes co-expressed across one or more conditions (such as cell types) that we call gene modules or latent variables/LVs.
\textbf{b)} The integration process in PhenoPLIER uses low-dimensional representations (matrices \(\mathbf{Z}\) and \(\mathbf{B}\)) learned from large gene expression datasets (top).
We used gene-drug information \(\mathbf{L}\) from LINCS L1000 and gene-trait associations \(\mathbf{M}\) from TWAS: PhenomeXcan was used as the discovery cohort, and eMERGE as replication (middle).
PhenoPLIER provides three computational components (bottom):
1) an LV-based regression model that associates an LV \(j\) (\(\mathbf{Z}_j\)) with a trait \(i\) (\(\mathbf{M}_i\)),
2) a clustering framework that learns groups of traits from TWAS associations projected into the LV space (\(\hat{\mathbf{M}}\)),
and 3) an LV-based drug repurposing approach that uses the projection of TWAS (\(\hat{\mathbf{M}}\)) and LINCS L1000 (\(\hat{\mathbf{L}}\)) into the LV space.
\textbf{c)} Genes that are part of LV603, termed as a neutrophil signature {[}\protect\hyperlink{ref-14rnBunuZ}{44}{]}, were expressed in relevant cell types (top).
LV603 was associated in PhenoPLIER with neutrophil counts and other white blood cells (bottom, showing the top 10 traits for LV603 after projecting gene-trait associations in PhenomeXcan).
eQTLs: expression quantitative trait loci;
MVN: multivariate normal distribution;
PBMC: peripheral blood mononuclear cells;
mDCs: myeloid dendritic cells.}\label{fig:entire_process}
}
\end{figure}

PhenoPLIER is a flexible computational framework that combines gene-trait and gene-drug associations with gene modules expressed in specific contexts (Figure \ref{fig:entire_process}a).
The approach uses a latent representation (with latent variables or LVs representing gene modules) derived from a large gene expression compendium (Figure \ref{fig:entire_process}b, top) to integrate TWAS with drug-induced transcriptional responses (Figure \ref{fig:entire_process}b, middle) for a joint analysis.
The approach consists in three main components (Figure \ref{fig:entire_process}b, bottom, see \protect\hyperlink{sec:methods}{Methods}):
1) an LV-based regression model to compute an association between an LV and a trait,
2) a clustering framework to learn groups of traits with shared transcriptomic properties,
and 3) an LV-based drug repurposing approach that links diseases to potential treatments.
We performed extensive simulations for our regression model (\protect\hyperlink{sm:reg:null_sim}{Supplementary Note 1}) and clustering framework (\protect\hyperlink{sm:clustering:null_sim}{Supplementary Note 2}) to ensure proper calibration and expected results under a model of no association.

@@

\begin{equation}
\begin{split}
Cor(\mathbf{P}_{i}, \mathbf{P}_{j}) & = Cor(\mathbf{T}_{i} \mathbf{V}_{i}^{\top} \mathrm{diag}(\lambda_i)^{-1/2}, \mathbf{T}_{j} \mathbf{V}_{j}^{\top} \mathrm{diag}(\lambda_j)^{-1/2}) \\
& = \mathrm{diag}(\lambda_i)^{-1/2} \mathbf{V}_{i} (\frac{\mathbf{T}_{i}^{\top} \mathbf{T}_{j}}{n-1}) \mathbf{V}_{j}^{\top} \mathrm{diag}(\lambda_j)^{-1/2},
\end{split}
\label{eq:reg:cor_pp}\end{equation}

@@

\begin{equation}
\begin{split}
\frac{(\mathbf{T}_{i}^{\top} \mathbf{T}_{j})_{kl}}{n-1} & = Cor(\mathbf{t}_k^i, \mathbf{t}_l^j) \\
 & = \frac{ Cov(\mathbf{t}_k, \mathbf{t}_l) } { \sqrt{\widehat{\mathrm{var}}(\mathbf{t}_k) \widehat{\mathrm{var}}(\mathbf{t}_l)} } \\
 & = \frac{ Cov(\sum_{a \in \mathrm{model}_k} w_a^k X_a, \sum_{b \in \mathrm{model}_l} w_b^l X_b) }  {\sqrt{\widehat{\mathrm{var}}(\mathbf{t}_k) \widehat{\mathrm{var}}(\mathbf{t}_l)} } \\
 & = \frac{ \splitfrac{\sum_{a \in \mathrm{model}_k}{b \in \mathrm{model}_l}} w_a^k w_b^l Cov(X_a, X_b)} {\sqrt{\widehat{\mathrm{var}}(\mathbf{t}_k) \widehat{\mathrm{var}}(\mathbf{t}_l)} } \\
        & = \frac{ \splitfrac{\sum_{a \in \mathrm{model}_k}{b \in \mathrm{model}_l}} w_a^k w_b^l \Gamma_{ab}} {\sqrt{\widehat{\mathrm{var}}(\mathbf{t}_k) \widehat{\mathrm{var}}(\mathbf{t}_l)} },
\end{split}
\label{eq:reg:corr_genes}\end{equation}

@@

Note that, since we used the MultiXcan regression model (Equation (\ref{eq:multixcan})), \(\mathbf{R}\) is only an approximation of gene correlations in S-MultiXcan.
As explained before, S-MultiXcan approximates the joint regression parameters in MultiXcan using the marginal regression estimates from S-PrediXcan in (\ref{eq:spredixcan}) with some simplifying assumptions and different genotype covariance matrices.
This complicates the derivation of an S-MultiXcan-specific solution to compute \(\mathbf{R}\).
To account for this, we used a submatrix \(\mathbf{R}_{\ell}\) corresponding to genes that are part of LV \(\ell\) only (top 1\% of genes) instead of the entire matrix \(\mathbf{R}\).
This simplification is conservative since correlations are accounted for top genes only.
Our simulations (\protect\hyperlink{sm:reg:null_sim}{Supplementary Note 1}) show that the model is approximately well-calibrated and can correct for LVs with adjacent and highly correlated genes at the top (e.g., Figure \ref{fig:reg:nulls:qqplot:lv234}).
The simulation also identified 127 LVs in which the model was not well-calibrated (e.g., Figure \ref{fig:reg:nulls:qqplot:lv914}).
As this can be attributed to limitations in accurately computing a gene correlation matrix, we excluded these LVs from our main analyses.

In Equation (\ref{eq:reg:corr_genes}), for each gene, we only considered tissue models present in S-PrediXcan results, as well as SNPs present in GWAS used as input for the TWAS approaches.
This is necessary to obtain more accurate correlation estimates {[}\protect\hyperlink{ref-1FFzCXo1s}{24}{]}.
Therefore, we computed different correlation matrices for PhenomeXcan and eMERGE.
In PhenomeXcan, most of the GWAS (4,049) were obtained from the UK Biobank using the same pipeline and including the same set of SNPs, so a single correlation matrix was used for this set.
For the rest, we used a single correlation matrix for each group of traits that shared the same or most of the SNPs.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.6954}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0993}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0993}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1060}}@{}}
\caption{Significant trait associations of LV246 in PhenomeXcan.
\label{tbl:sup:phenomexcan_assocs:lv246}}\label{tbl:sup:phenomexcan_assocs:lv246}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cases
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
FDR
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Cases
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
FDR
\end{minipage} \\
\midrule()
\endhead
Triglycerides NMR & 21,559 & & 1.66e‑26 \\
LDL Cholesterol NMR & 13,527 & & 3.92e‑26 \\
High cholesterol (self-reported) & 361,141 & 43,957 & 1.08e‑24 \\
Cholesterol lowering medication & 193,148 & 24,247 & 4.28e‑24 \\
Treatment/medication code: simvastatin & 361,141 & 40,921 & 2.56e‑19 \\
CH2DB NMR & 24,154 & & 1.05e‑15 \\
Cholesterol lowering medication & 165,340 & 38,057 & 9.58e‑15 \\
Treatment/medication code: atorvastatin & 361,141 & 10,805 & 2.54e‑14 \\
Illnesses of mother: Alzheimer's disease/dementia & 331,041 & 28,507 & 2.76e‑08 \\
Illnesses of father: Alzheimer's disease/dementia & 312,666 & 15,022 & 2.76e‑08 \\
Alzheimers Disease & 54,162 & 17,008 & 1.10e‑07 \\
Non-butter spread type details: Flora Pro-Active or Benecol & 190,094 & 29,048 & 5.63e‑07 \\
Illnesses of siblings: Alzheimer's disease/dementia & 279,062 & 1,609 & 6.16e‑07 \\
Any dementia & 361,194 & 243 & 2.86e‑05 \\
Illnesses of father: None of the above (group 1) & 314,797 & 116,736 & 3.56e‑05 \\
Medication for cholesterol, blood pressure, diabetes, or take exogenous hormones (females) & 193,148 & 133,338 & 1.10e‑04 \\
Treatment/medication code: lipitor 10mg tablet & 361,141 & 2,584 & 1.55e‑04 \\
Treatment/medication code: rosuvastatin & 361,141 & 2,227 & 1.37e‑03 \\
Illnesses of father: Heart disease & 318,570 & 104,110 & 1.89e‑03 \\
Dementia & 361,194 & 157 & 9.58e‑03 \\
Mother still alive & 355,029 & 140,246 & 1.76e‑02 \\
Job SOC coding: Librarians & 91,149 & 1,248 & 3.22e‑02 \\
Alzheimer's disease & 361,194 & 119 & 3.61e‑02 \\
\bottomrule()
\end{longtable}
